Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-18 paper/paper.tex
Changes

Any polynomial functor can be represented as a container (see accompanying code
for a mechanisation).  For example, a container that is equivalent to
$F\; X = 1 + A \times X$ requires a shape that can distinguish two cases,
indicating that there are two constructors to build an object of type $F\;X$,
e.g. $S_{F} = 1 + A$. We need to consider two cases for the family of positions:
$P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
left case; and $P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence
of $X$ in the right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

Note however, that the type $\forall s : S.\; P\;s \to X$ would force us to add
the functional extensionality axiom to deal with the heterogeneous equality of
positions. Thus, we require that all the types used in our definitions are
\emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
morphisms need to be \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
tactics to automatically discharge proofs that morphisms are proper,  for some
simple cases.

We define containers in terms of shapes, \emph{base} positions (\coq{APos : Set}),
and \emph{decidable validity predicates} that
specify when a base position is valid in a shape. Since validity predicates are in
\coq{Prop}, they will be erased during extraction:
\begin{coqcode}
Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
\end{coqcode}

Given two \coq{x y : App F X}, we define
\coq{x =e y} as the setoid equality of the shapes as
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}

>>>>

Any polynomial functor can be represented as a container (see accompanying
code).  For example, a container that is equivalent to $F\; X = 1 + A \times X$
requires a shape that can distinguish two cases, indicating that there are two
constructors to build an object of type $F\;X$, e.g. $S_{F} = 1 + A$. We need to
consider two cases for the family of positions:
$P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
left case; and $P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence
of $X$ in the right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

The type $\forall s : S.\; P\;s \to X$ would force us to add
the functional extensionality axiom. Thus, we require that all the types used in
our definitions are \emph{setoids}, where we use \coq{x =e y} to represent
setoid equality. All morphisms need to be \emph{proper} morphisms that respect
setoid equalities. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
tactics to automatically discharge proofs that morphisms are proper,  for some
simple cases.

%% WARNING: removing the point about axiom K leads to leaving the use of
%% decidable validity predicates unexplained -- do not remove!
We define containers in terms of shapes, \emph{base} positions
(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
base position is valid in a shape. Since validity predicates are in \coq{Prop},
they will be erased during extraction, and since they are decidable, they
prevent the use of axiom K to deal with heterogeneous equalities of positions:
\begin{coqcode}
Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
\end{coqcode}

%% Make sure you do not cut text so much that it ends up being inconsistent with
%% the code! I just restored an old explanation and cut some of it in a consistent way
We define the setoid equality of two container extensions \coq{x y : App F X} as
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}
